The Ranked Nearest Neighbour Interchange space

### 2019-06-28 paper.tex
\mrca, \rank, \parent

\newcommand{\np}{\mathcal{NP}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\rspr}{\mathrm{rSPR}}

@@

In this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(T, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|\rank(v) - \rank(u)| = l$.
Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
We also say that $v$ \emph{induces} the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
	\label{fig:RNNI}
\end{figure}

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster node $v$ in $R$ with $rank(v) = i$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

\todo{This text comes out of the blue.
Generally, before going into explanations like this it's a good idea to say what's ahead.}
One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
\todo{This should have been stated before, where you say what is going on.}
Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p|_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p|_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta := |rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    %
    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
    % \end{enumerate}
\end{proof}

@@

Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
\todo{Pseudo-Code for algorithm?}

@@

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
    We also know $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = n-1$
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.

@@

% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.

@@

% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.

@@

% 	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.

>>>>

\newcommand{\np}{\mathcal{NP}}

\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\rspr}{\mathrm{rSPR}}

@@

In this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(T, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$ or $\rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|\rank(v) - \rank(u)| = l$.
Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
We also say that $v$ \emph{induces} the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the \emph{parent} of $x$ and will be denoted by $\parent(x)$.
We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $\mrca(\{1,2\})$ and $\mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
	\label{fig:RNNI}
\end{figure}

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster node $v$ in $R$ with $\rank(v) = i$
		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

\todo{This text comes out of the blue.
Generally, before going into explanations like this it's a good idea to say what's ahead.}
One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $\mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
\todo{This should have been stated before, where you say what is going on.}
Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $\mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

\begin{lemma}
    Let $T$ and $R$ be two trees on $n+1$ taxa.
    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
    Then it is $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    \label{lemma:distance_delete_taxon}
\end{lemma}

\begin{proof}
    Let $p$ be a shortest path from tree $T$ to $R$ and $p|_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
    This new path $p|_n$ is then a path from $T$ to $R$.
    Showing that at least $\delta := |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
    By the following two observations one can easily see that this claim is true, indeed:
    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    % \end{enumerate}
\end{proof}

@@

Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
\todo{Pseudo-Code for algorithm?}

@@

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.

@@

% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $\rank(int_x) < \rank(int_y)$ as in Figure~\ref{fig:nni_moves}.

@@

% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $\rank(int_x) > \rank(int_y)$.

@@

% 	Then the labels of the two internal nodes change such that $\rank(int_x) > \rank(int_y)$.
